---
title: "HW4"
author: "Liam Knight"
date: "2026-03-06"
output: html_document
editor_options: 
  markdown: 
    wrap: sentence
---

```{r, include=FALSE}
library(plotly)
library(tidyverse)
```

Recall the histogram for `cut` categories from the `diamonds` data set:

```{r}
diamonds %>%
  plot_ly( x = ~cut, showlegend = FALSE ) %>%
  add_histogram(hoverinfo = "x+y") %>%
  dplyr::group_by( cut ) %>%
  summarise(n = n()) %>%
  add_text(
    text = ~scales::comma(n), y = ~n,
    textposition = "top middle",
    cliponaxis = FALSE,
    hoverinfo = "none"
  )
```

Please add to it the following changes:

1.  make the bar chart a stacked bar chart: so that the $x$-axis values are still according to the variable `cut`, but the internal division of each bar is according to the `clarity` variable.
2.  change the colors of the bar segments to be in accordance with the `"Blues"` palette.
3.  make sure to have the tooltip content for each bar segment as it is in the below plot (“The cut is … such diamonds!”)
4.  set the color of the font in the tooltip to red (hint: `hoverlabel`).
5.  set the background of the tooltip to yellow.
6.  set the background color of the entire plot to `"pink"`.
7.  set the background color of the plotting area to `"lightgray"`.
8.  set the borders of the bars to black and thickness (`span`) of 1.
9.  set the color of the x-axis labels to `"green"`.
10. set the title of the x-axis to `"Cut"` (instead of “cut”).
11. set the color of the x-axis title to `"darkred"` and its font to `"Times New Roman"`.
12. change the color of the y-axis labels to `"darkorange"` (be careful not to change other things with it).
13. remove the y-axis title.
14. set the left margin of the plot to 25px.
15. change the color of the text values to `"red"`. No tooltip is associated with these values.
16. add the title `"Ugly, ugly plot"` in the `"Old Standard TT"` font at the top left corner.

**Solution:**

```{r}
diamonds %>%
  select(cut, clarity) %>%
  plot_ly(x = ~cut, showlegend = FALSE, colors = "Blues") %>%
  add_histogram(color = ~clarity, 
                marker = list(line = list(color = "black", 
                                          width = 1)
                              ),
                hoverlabel = list(bgcolor = "yellow",
                                  font = list(color = "red")
                                  ),
                hovertemplate = "The cut is %{x},\n the clarity is %{fullData.name},\n and there are %{y} such diamonds!<extra></extra>"
                ) %>%
  layout(barmode = "stack", 
         margin = list(l = 25),
         title = list(text = "Ugly, ugly plot", 
                      font = list(family = "Old Standard TT"), 
                      x = 0
                      ),
         paper_bgcolor = "pink",
         plot_bgcolor = "lightgray",
         xaxis = list(title = list(text ="Cut", 
                                   font = list(color = "darkred",
                                               family = "Times New Roman")
                                   ),
                      tickfont = list(color = "green")
                      ),
         yaxis = list(title = "",
                      tickfont = list(color = "darkorange")
                      )
         ) %>%
  dplyr::group_by(cut) %>%
  summarise(n = n()) %>%
  add_text(
    text = ~scales::comma(n), y = ~n,
    textfont = list(color = "red"),
    textposition = "top middle",
    cliponaxis = FALSE,
    hoverinfo = "none"
  )

```
