---
title: "HW 3"
author: "Liam Knight"
date: "2026-02-18"
output: html_document
---

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

packages <- c('knitr', 
              'tidyverse', 
              'janitor',
              'plotly',
              'shiny',
              'gapminder',
              'rsconnect')

lapply(packages, library, character.only=TRUE)
```

Link to Shiny app in Question 3: https://liamknight.shinyapps.io/gapminderapp/ 

# Question 1

Create a graph similar to and based on the work you did in Homework 2, Question 1. This time, depict on the y-axis the life expectancy of each country relative to the life expectancy in the US (for every given year). That is, the y-axis will present data like 

`(life expectancy for Cuba in 1919) / (life expectancy for the US in 1919)`, 
`(life expectancy for Cuba in 1920) / (life expectancy for the US in 1920)`, 
`(life expectancy for Iraq in 1919) / (life expectancy for the US in 1919)`, 
`(life expectancy for Iraq in 1920) / (life expectancy for the US in 1920)`, etc.

Hint: if you do this correctly, the point that represents the US should always have a y-value of 1. Add a layer to the plot that highlights the marker for the US throughout the animated sequence (in any reasonable way you choose). Add a layer to the plot that shows a horizontal line at the y-value of 1.

**Question 1 Solution:**

```{r}
#read in data
recreated_gapminder <- 
  read.csv(
   "data/recreated_gapminder.csv", 
   header = TRUE, 
   stringsAsFactors = FALSE,
   check.names = FALSE
   ) 
  
#wrangle data
us_dat <- recreated_gapminder |>
  #use janitor to clean extra (unnamed) row index column
  clean_names() |>
  filter(country == 'United States') |>
  rename(us_life_exp = "life_exp") |>
  select(year, us_life_exp)

#join US life expectancies to main dataset 
dat <- recreated_gapminder |>
  clean_names() |>
  left_join(y = us_dat, by = 'year') |>
  mutate(prop = life_exp/us_life_exp) 
  
```


```{r, warning=FALSE}
#create plot
gg <- 
  ggplot(dat, aes(gdp_per_cap, prop, color = continent)) +
  geom_point(aes(size = pop, frame = year, ids = country)) +
  geom_point(
    data = filter(dat, country == "United States"),
    aes(x = gdp_per_cap, y = prop, size = pop, frame = year, ids = country),
    colour = "purple", fill = "purple"
    ) +
  scale_x_log10() + 
  theme(legend.title = element_blank(),
        plot.title = element_text(size = 12, hjust = 0.5, margin = margin(b = 2)),) +
  labs(
    title = "Life expectancies as relative to the United States vs GDP per capita \n(United States in purple)", 
    x = "GDP per capita",
    y = "Life expectancy relative to the US"
    ) +
  geom_hline(yintercept = 1)

ggplotly(gg)

```


# Question 2

(no question 2)


# Question 3

Complete our gapminder app so that it uses ggplot to show a similar plot to the gapminder plot from homework 1 question 2.1 (same as in the introduction):

- Data is filtered first according to year
- Then, data is filtered according to the continent
- Lastly, (a feature we add here to the original functionality) also filter the data according to quantiles of interest in income (so that if the range is 0-50, we’ll see the lower half of countries for a given year)
- The plot should be based on ggplot (not using ggplotly here).
- You may use the small gapminder dataset as demonstrated in class or utilize the larger one you prepared in the second homework assignment.

In the HTML you submit, please comment on the meaning of the additional filtering by income quantiles in step 3. One way to help clarify to yourself this meaning may be performing the filtrations in a different order. How do the results differ? What kind of insights are can be made by the different methods? In order to exemplify your arguments, add to your answer individual interactive plots created with ggplotly for specific years (no animation) where the data are filtered by quantiles of interest and year before being sent to ggplot (i.e., you don’t need to add functionality to the interactive plot, use it on data that you manipulate in advance).

**Question 3 Solution:**


Link to Shiny app:
https://liamknight.shinyapps.io/gapminderapp/

```{r}


df <- gapminder::gapminder

ui <- fluidPage(
  titlePanel("Gapminder"),
  sidebarLayout(
    sidebarPanel(
      "Interactive plotting of gapminder data using R shiny",
      #checklist for "Choose a continent"
      checkboxGroupInput(
        inputId = "continent",
        label = "Choose a continent",
        choices = levels(df$continent),
        selected = levels(df$continent)
      ),
      #slider for "Income Percentiles of interest"
      sliderInput(
        inputId = "percentile",
        label = "Income percentiles of interest",
        min = 0,
        max = 100,
        value = c(0,100)
      ),
      #slider for "Year"
      sliderInput(
        inputId = "year",
        label = "Year",
        min = 1952,
        max = 2007,
        value = 1952,
        step = 5,
        sep = "",
        animate = animationOptions(interval = 500, loop = FALSE)
      )
    ),
    mainPanel(
      plotOutput(outputId = "gapminder_plot")
    )
  )
)

server <- function(input, output) {
  output$gapminder_plot <- renderPlot({
    #store input selections
    selected_year <- input$year
    selected_continents <- input$continent    
    selected_percentile <- input$percentile

    df |>
      #group by year so that percentiles are recalculated based on the dataset
      #for each year, rather than GDPs for 1952 being compared against
      #GDPs for 2007, for example
      group_by(year) |>
      #create new variable for percentile
      mutate(percentile = 100 * percent_rank(gdpPercap)) |>
      ungroup() |>
      #filter by input selections
      filter(year == selected_year) |>          
      filter(continent %in% selected_continents) |> 
      filter(percentile <= selected_percentile) |>
      ggplot(aes(gdpPercap, lifeExp, color = continent)) +
      scale_x_log10(
        limits = c(1e2, 1e5),
        breaks = c(100, 1000, 10000, 100000),
        labels = comma) +
      scale_size_continuous(
        name = "Population size", 
        range = c(1, 8), 
        labels = comma) +
      scale_color_discrete(name = "Continent") +
      geom_point(aes(size = pop)) +
      ylim(0, 80) +
      labs(x = "GDP per capita",
           y = "Life expectancy (years)",
           title = "Global life expectancies as a function of gross domestic product (GDP) per capita",
           subtitle = "1952-2007",
           color = "Continent",
           size = "Population size")
  })
}

shinyApp(ui, server)
```

    
Below, we manipulate our data by defining that we will filter only for the continents of Asia and Africa. However, in one plot, we have filtered for continent prior to calculating percentiles and filtering on those percentiles, and in the other plot, we calculate the percentiles and filter by them first, then filter by continent. Note that the two plots look different. 

When we calculate the percentiles AFTER filtering by continent, the percentiles are calculated on a different dataset: they are calculated only for the data for the countries in Africa and Asia, rather than on the entire global set of countries. This means that the percentiles reflect only where a given value lies in terms of quantile rank for the set of countries in Asia and Africa in that year. Therefore, we are seeing insights about where a country ranks within those two continents.

On the other hand, when we calculate the percentiles FIRST, then filter by continent second, we end up seeing countries that fell in the 25th percentile globally for 1952, rather than only countries that fell in the 25th percentile in Africa and Asia that year. Therefore, we see more countries on the plot, and we are seeing insights about where a country ranks within the entire global set of countries across all continents.


```{r}
#Individual interactive plots

#manipulate selecting continents
sel_continents <- c("Africa", "Asia")

#year 1952, only countries within the 25th percentile, filtering by continent BEFORE percentile
df_1952_25_1 <- df |>
  #filter
  filter(year == 1952) |>
  filter(continent %in% sel_continents) |>
  mutate(percentile = 100 * percent_rank(gdpPercap)) |>
  filter(percentile <= 25) 
  
gg_1952_25_1 <- ggplot(df_1952_25_1, aes(gdpPercap, lifeExp, color = continent)) +
  scale_x_log10(limits = c(1e2, 1e5)) +
  geom_point(aes(size = pop)) +
  ylim(0, 80) +
  theme(legend.title = element_blank()) +
  labs(title = "1952, 25th percentile, filter for continent BEFORE percentile")

ggplotly(gg_1952_25_1)



#year 1952, only countries within the 25th percentile, filtering by continent AFTER percentile
df_1952_25_2 <- df |>
  filter(year == 1952) |>
  mutate(percentile = 100 * percent_rank(gdpPercap)) |>
  filter(percentile <= 25) |>
  filter(continent %in% sel_continents)
  
gg_1952_25_2 <- ggplot(df_1952_25_2, aes(gdpPercap, lifeExp, color = continent)) +
  scale_x_log10(limits = c(1e2, 1e5)) +
  geom_point(aes(size = pop)) +
  ylim(0, 80) +
  theme(legend.title = element_blank()) + 
  labs(title = "1952, 25th percentile, filter for continent AFTER percentile")

ggplotly(gg_1952_25_2)



```

